---
output: pdf_document
classoption: a4paper
---

```{r echo = FALSE}
library(knitr)
opts_knit$set(root.dir = '..')
```

# Data Management

## Data sources

Data was sourced from public sources, mostly from the [Department of Education's open data portal][1]. The complete list of data inputs is as follows:

```{r, echo = FALSE, message = FALSE}
library(readxl)
library(pander) 
panderOptions('table.split.table', Inf)
set.caption('Data descriptions and sources')
pander(read_excel("Documentation/Data Descriptions.xlsx"))
```

[1]: http://www.deped.gov.ph/datasets

## Data processing

Before analysis, we process the various datasets as follows:

* Grade levels in the enrollment data were harmonized across the years (e.g. Grade 7 and Year 1 are combined, and so on).
* Locations were integrated into the masterlist of schools, when available.
* Teachers, rooms, and budget data were integrated into the masterlist of schools, when available.
* Data were saved into RData images for easy access.